\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts{}

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Parallel Voxelization of Triangle Meshes}

\author{\IEEEauthorblockN{1\textsuperscript{st} Jose Francisco Paca Sotero}
    \IEEEauthorblockA{\textit{Faculty of Computing} \\
        \textit{University of Engineering and Technology (UTEC)}\\
        Lima, Peru \\
        jose.paca@utec.edu.pe}
    \and
    \IEEEauthorblockN{2\textsuperscript{nd} Diva Stewart Maquera Bobadilla}
    \IEEEauthorblockA{\textit{Faculty of Computing} \\
        \textit{University of Engineering and Technology (UTEC)}\\
        Lima, Peru \\
        diva.maquera@utec.edu.pe}
    \and
    \IEEEauthorblockN{3\textsuperscript{rd} Daniel Ignacio Casquino Paz}
    \IEEEauthorblockA{\textit{Faculty of Computing} \\
        \textit{University of Engineering and Technology (UTEC)}\\
        Lima, Peru \\
        daniel.casquino@utec.edu.pe}
}

\maketitle

\begin{abstract}
    This paper presents a parallel implementation of conservative voxelization for
    triangle meshes. The main implementation uses MPI to distribute triangles across
    processes, build private bit-packed voxel grids, and merge the final result with
    a bitwise OR reduction. To justify the selected paradigm, two shared-memory
    OpenMP variants are also implemented: one with a private grid per thread and one
    with a shared grid updated through atomic OR operations. The evaluation compares execution
    time, speedup, efficiency, and experimental overhead for small and moderate grid
    resolutions.
\end{abstract}

\begin{IEEEkeywords}
    voxelization, MPI, OpenMP, parallel computing, overhead, PRAM
\end{IEEEkeywords}

\section{Introduction}

Voxelization is the process of discretizing a 3D object into an axis-aligned
grid of volumetric pixels. This representation is used in computer graphics,
medical imaging, urban analysis, simulation, and interactive applications. This
project studies triangle-mesh voxelization as a parallel computing problem:
given a set of triangles and a fixed grid resolution, mark the voxels occupied
by the mesh.

The first version of the project modeled the algorithm in PRAM and identified
task parallelism over triangles. This final version keeps that idea but
implements and measures three concrete CPU variants: MPI, OpenMP with private
grids, and OpenMP with atomic updates. CUDA and tile-based voxelization are
discussed as future work because they require additional GPU memory management
or spatial indexing to be correct and measurable.

\section{Algorithm}

\subsection{Input and Representation}

The input is a triangle mesh $T=\{t_0,t_1,\ldots,t_{m-1}\}$ and a grid
resolution $r$. The voxel grid has $V=r^3$ cells. To reduce memory, the grid is
stored as a bitset implemented with 64-bit words, so the number of stored words
is $w=\lceil V/64\rceil$.

For each triangle, the current implementation computes its axis-aligned
bounding box (AABB) in grid coordinates and marks every voxel inside that box.
This is conservative: it can mark false positives, but it is simple,
deterministic, and exposes parallel work clearly.

\subsection{PRAM Model}

The high-level algorithm can be modeled as CRCW PRAM because different workers
may write to the same voxel. The write conflict is benign when the operation is
a logical OR, as marking the same voxel multiple times still leaves the final
bit set to one.

\begin{algorithm}
    \caption{Voxelization (PRAM)}
    \begin{algorithmic}[1]
        \State$B \gets \text{empty bitset}$
        \State$T \gets \text{triangulate}(\text{model.load}(path)).faces$
        \ForAll{$t$ \textbf{in} $T$ \textbf{pardo}}
        \ForAll{candidate voxel $v$ \textbf{inside AABB of} $t$}
        \State$i \gets \text{index}(v)$
        \State$B[i] \gets B[i] \lor 1$
        \EndFor{}
        \EndFor{}
        \State\Return$B$
    \end{algorithmic}
\end{algorithm}

$c_i$ is defined as the number of candidate voxels visited for triangle $t_i$. Then,
$C=\sum_{i=0}^{m-1} c_i$. The sequential work of the voxelization stage is:

\[
    T_s = \Theta(m + C)
\]

With unbounded PRAM processors over triangles, the span is dominated by the
largest triangle AABB:\@

\[
    T_{\infty} = O(\max_i c_i)
\]

Using Brent's theorem, an ideal parallel bound is then:

\[
    T_p = O\left(\frac{m+C}{p} + \max_i c_i\right)
\]

This bound explains the main limitation of triangle partitioning: if a few
triangles cover very large AABBs, load balancing can degrade.

\section{Parallel Implementations}

\subsection{MPI}

The MPI implementation is the main distributed-memory version. Rank zero loads
the mesh and distributes triangles with \texttt{MPI\_Scatterv}. Then, each rank
builds its own local bitset grid. The final grid is obtained with
\texttt{MPI\_Reduce} and \texttt{MPI\_BOR}.

\[
    T_p^{\text{MPI}} \approx \frac{C}{p} + T_{\text{scatter}}(m,p) + T_{\text{reduce}}(w,p)
\]

This design avoids data races during voxel marking, as each process writes to
private memory only. Its main overhead comes from process initialization,
triangle distribution, synchronization, and reduction of the bit-packed grid.

\subsection{OpenMP with Private Grids}

The \texttt{omp\_private} variant uses one process and multiple threads. Each
thread writes to a private grid, and all grids are merged with an OR reduction:

\[
    T_p^{\text{OMP-private}} \approx \frac{C}{p} + T_{\text{reduce\_mem}}(w,p)
\]

This version avoids atomic operation overhead, but it increases memory usage by
a factor that is close to the number of threads.

\subsection{OpenMP with Atomic Updates}

The \texttt{omp\_atomic} variant uses a shared bitset grid. Each update uses an
atomic OR on the 64-bit word that contains the target voxel bit:

\[
    T_p^{\text{OMP-atomic}} \approx \frac{C}{p} + A C_{\text{atomic}} + C_{\text{contention}}
\]

This version saves memory, but several threads can contend for the same 64-bit
word even when they update different voxel bits.

\section{Experimental Methodology}

The measured region is \texttt{voxel\_seconds}. For each mode, the baseline
$T_s$ is the same mode with $p=1$. The metrics are:

\begin{align}
    S(p)   & =\frac{T_s}{T_p},    &
    E(p)   & =\frac{S(p)}{p},       \\
    T_o(p) & =pT_p-T_s,           &
    R_O(p) & =\frac{T_o(p)}{T_s}.
\end{align}

Local experiments used the OBJ model \texttt{blocky\_creeper\_like}, grid
resolutions $32$, $64$, and $128$, $p\in\{1,2,4\}$, and three repetitions per
configuration. Table I reports the median for resolution $128$, where the
parallel work is large enough to expose speedup in some modes.

\section{Results}

\begin{table}[t]
    \centering
    \caption{Local results for resolution 128, median of three runs.}
    \begin{tabular}{lrrrrr}
        \toprule
        Mode        & $p$ & $T_p$ (s) & $S(p)$ & $E(p)$ & $O/T_s$ \\
        \midrule
        MPI         & 1   & 0.002703  & 1.000  & 1.000  & 0.000   \\
        MPI         & 2   & 0.001755  & 1.540  & 0.770  & 0.299   \\
        MPI         & 4   & 0.002288  & 1.181  & 0.295  & 2.386   \\
        OMP atomic  & 1   & 0.003745  & 1.000  & 1.000  & 0.000   \\
        OMP atomic  & 2   & 0.002837  & 1.320  & 0.660  & 0.515   \\
        OMP atomic  & 4   & 0.002061  & 1.817  & 0.454  & 1.201   \\
        OMP private & 1   & 0.003016  & 1.000  & 1.000  & 0.000   \\
        OMP private & 2   & 0.002748  & 1.098  & 0.549  & 0.822   \\
        OMP private & 4   & 0.002948  & 1.023  & 0.256  & 2.910   \\
        \bottomrule
    \end{tabular}
\end{table}

For resolutions $32$ and $64$, the measured times are very small and overhead
dominates. At resolution $128$, MPI obtains its best local point at $p=2$.
OpenMP atomic improves up to $p=4$ in this small mesh, while OpenMP private
does not clearly outperform the other variants because the private-grid
reduction is significant relative to the computation.

\section{Khipu Validation}

The project was also validated on UTEC's Khipu cluster using Slurm, OpenMPI,
and CMake. The code was adjusted so it can build without Assimp by using a
small OBJ fallback loader for benchmark models. A conservative debug job used
up to four CPUs and small resolutions. This run is treated as portability
validation, not as a definitive benchmark, because the measured times are in
the microsecond to millisecond range and are therefore sensitive to scheduler
and environment noise.

One operational issue was observed: the available OpenMPI environment did not
work correctly with direct \texttt{srun} execution for this program. The stable
approach was to reserve resources with Slurm and use \texttt{mpirun} inside the
allocation.

\section{Discussion}

The results show that increasing $p$ does not always reduce execution time. For
strong scaling, the problem size is fixed, so local computation decreases
approximately as $B_c/p$, but reduction, synchronization, launch, and memory
costs do not disappear. Therefore, the best $p$ is not necessarily the largest
available $p$.

MPI is defensible as the main implementation because it represents distributed
memory directly, avoids concurrent writes through private grids, and expresses
the final merge as a natural bitwise OR reduction. OpenMP is simpler on one
machine and can have lower overhead, but its variants expose different costs:
private grids spend memory and reduction bandwidth, while atomic updates can
suffer contention.

The main weakness of triangle-based partitioning is load balance. If a mesh has
only a few large triangles, some workers may receive much more work than
others. A tile-based algorithm could assign disjoint voxel regions to workers
and eliminate the final OR reduction, but a naive tile implementation would
compare too many triangles against too many voxels. Its work is
$\Theta(mr^3/p)$ per parallel step without spatial filtering, while the current
triangle-based implementation visits only $C=\sum_i c_i$ candidate voxels.
For example, a mesh with $69{,}451$ triangles at $r=512$ would imply about
$9.3\times10^{12}$ triangle-voxel tests in the naive all-triangles-per-tile
approach. Practical tile-based voxelizers need spatial filtering structures
such as binning, BVH, octrees, or sparse grids.

\subsection{Why CUDA Was Not the Main Delivery}

CUDA is technically attractive, especially when a local GPU is available, but
the limiting issue is not only whether a kernel can be launched. A correct GPU
version needs a memory layout, atomic or block-local reductions for bit updates,
host-device transfer timing, and output validation against the CPU
implementation.

The memory cost also grows quickly for dense grids. A bit-packed grid requires
$r^3/8$ bytes, while a byte-per-voxel grid requires $r^3$ bytes. At $r=4096$,
the bit-packed grid already needs 8 GiB, and a byte grid needs 64 GiB. If each
GPU block or worker stores four private bit-packed grids before reducing, the
same $r=4096$ case needs 32 GiB, above a 12 GiB local GPU. With 12 GiB of VRAM,
the theoretical maximum dense bit-packed grid is about $r=4688$ for a single
grid, about $r=2953$ with four private grids, and about $r=2344$ with eight
private grids, before counting triangles, auxiliary buffers, CUDA runtime
overhead, and temporary reductions.

Therefore, a CUDA implementation is reasonable future work only if paired with
a carefully chosen memory strategy and a validation plan. For this delivery,
MPI and OpenMP provide a simpler and more defensible comparison of distributed
and shared memory overheads.

\section{Conclusion}

The final implementation keeps MPI as the main solution and adds OpenMP
variants for comparison. The experiments confirm that overhead is central to
the analysis: for small grids, parallelism can lose to overhead, while for
larger fixed grids some configurations obtain speedup. Future work should
replace the AABB marking approximation with a triangle-box test, implement
tile-based partitioning with spatial filtering, and evaluate CUDA only when GPU
execution can be validated against the CPU outputs.

\section*{Use of AI}

OpenAI ChatGPT/Codex was used as an assistant to organize the report, review
the complexity derivation, compare MPI/OpenMP/CUDA tradeoffs, and prepare
documentation. The code, formulas, and conclusions were checked against the
repository implementation, official MPI/OpenMP/Khipu documentation, and the
technical references below. AI was not used as a primary source for
experimental results.

\begin{thebibliography}{00}
    \bibitem{b1} M. Schwarz and H.-P. Seidel, ``Fast parallel surface and solid voxelization on GPUs,'' \textit{ACM Transactions on Graphics}, vol. 29, no. 6, Article 179, 2010.
    \bibitem{b2} T. Akenine-Moller, ``Fast 3D triangle-box overlap testing,'' \textit{Journal of Graphics Tools}, vol. 6, no. 1, pp. 29--33, 2001.
    \bibitem{b3} I. Foster, \textit{Designing and Building Parallel Programs: Concepts and Tools for Parallel Software Engineering}. Addison-Wesley, 1995.
    \bibitem{b4} Open MPI Project, ``MPI\_Scatterv(3), MPI\_Bcast(3), and MPI\_Reduce(3) manual pages,'' Open MPI Documentation. [Online]. Available: https://www.open-mpi.org/doc/
    \bibitem{b5} OpenMP Architecture Review Board, ``OpenMP API Specification,'' 2024. [Online]. Available: https://www.openmp.org/specifications/
    \bibitem{b6} Open Asset Import Library, ``Assimp: Open Asset Import Library.'' [Online]. Available: https://www.assimp.org/
    \bibitem{b7} UTEC Khipu, ``Khipu documentation.'' [Online]. Available: https://docs.khipu.utec.edu.pe/
    \bibitem{b8} OpenAI, ``ChatGPT/Codex,'' large language model, 2026. [Online]. Available: https://chat.openai.com/
\end{thebibliography}

\end{document}
